\documentclass[11pt]{article}

\usepackage{amsmath, amssymb, amsfonts}
\usepackage{geometry}
\usepackage{enumitem}

\geometry{margin=1in}

\title{Security-Constrained Economic Dispatch (SCED) Formulation}
\author{}
\date{}

\begin{document}

\maketitle

\section*{Given}

We consider a power network operated under a DC power flow approximation.
Generation must be dispatched to meet demand at minimum cost while satisfying:

\begin{itemize}[noitemsep]
    \item System-wide power balance,
    \item Generator operating limits,
    \item Transmission thermal limits,
    \item $N$-$1$ security constraints (contingency feasibility).
\end{itemize}

The dispatch must remain feasible under all specified contingencies without post-contingency redispatch.

\section*{Parameters}

\begin{itemize}[noitemsep]
    \item $\mathcal{N}$ : Set of buses (nodes)
    \item $\mathcal{L}$ : Set of transmission lines
    \item $\mathcal{G}$ : Set of generators
    \item $\mathcal{D}$ : Set of loads
    \item $\mathcal{K} \subseteq \mathcal{L}$ : Set of contingencies
    \item $c_g$ : Marginal cost of generator $g$
    \item $g_g^{\min}, g_g^{\max}$ : Generation limits
    \item $d_d^{\text{nom}}$ : Nominal demand
    \item $F_l^{\max}$ : Thermal flow limit of line $l$
    \item $PTDF_{l,s}$ : Shift factor of line $l$ with respect to node $s$
    \item $PTDF_{l,s}^{(k)}$ : Shift factor under contingency $k$
    \item $n(g)$ : Node of generator $g$
    \item $n(d)$ : Node of load $d$
\end{itemize}

\section*{LP Variables}

\begin{itemize}[noitemsep]
    \item $g_g \ge 0$ : Power produced by generator $g$
    \item $d_d \ge 0$ : Power served to load $d$
\end{itemize}

 
These variables determine the economic dispatch. Generation variables determine supply, and demand variables allow the model to enforce power balance and load limits.

\section*{LP Objective Function}

\begin{equation}
\min_{g,d} \quad \sum_{g \in \mathcal{G}} c_g g_g
\end{equation}

 
The objective minimizes total generation cost, ensuring the most economical dispatch that satisfies reliability constraints.

\section*{LP Constraints}

\subsection*{1. System Power Balance}

\begin{equation}
\sum_{g \in \mathcal{G}} g_g
-
\sum_{d \in \mathcal{D}} d_d
= 0
\end{equation}

 
Ensures total generation equals total demand, maintaining physical feasibility of the power system.

\subsection*{2. Generator Capacity Limits}

\begin{equation}
g_g^{\min} \le g_g \le g_g^{\max}
\quad \forall g \in \mathcal{G}
\end{equation}

 
Generators cannot produce below minimum stable output or exceed technical capacity limits.

\subsection*{3. Load Limits}

\begin{equation}
0 \le d_d \le d_d^{\text{nom}}
\quad \forall d \in \mathcal{D}
\end{equation}

 
Prevents serving more than the available demand and ensures non-negativity of consumption.

\subsection*{4. Net Injection Definition}

For each node $s \in \mathcal{N}$:

\begin{equation}
inj_s =
\sum_{g: n(g)=s} g_g
-
\sum_{d: n(d)=s} d_d
\end{equation}

 
Defines nodal net injection used to compute transmission flows via PTDFs.

\subsection*{5. Base-Case Transmission Constraints}

\begin{equation}
- F_l^{\max}
\le
\sum_{s \in \mathcal{N}} PTDF_{l,s} \, inj_s
\le
F_l^{\max}
\quad \forall l \in \mathcal{L}
\end{equation}

 
Ensures no transmission line exceeds its thermal rating under normal operating conditions.

\subsection*{6. Contingency Constraints ($N$-$1$ Security)}

\begin{equation}
- F_l^{\max}
\le
\sum_{s \in \mathcal{N}} PTDF_{l,s}^{(k)} \, inj_s
\le
F_l^{\max}
\quad \forall l \in \mathcal{L}, \forall k \in \mathcal{K}
\end{equation}

 
Guarantees that the chosen dispatch remains feasible if any single contingency occurs.  
The same generation dispatch must satisfy all contingency flow limits (no corrective redispatch).

\newpage

\section*{Full SCED Formulation}

\begin{align}
\min_{g,d} \quad
& \sum_{g \in \mathcal{G}} c_g g_g
\\[6pt]
\text{s.t.} \quad
& \sum_{g \in \mathcal{G}} g_g
-
\sum_{d \in \mathcal{D}} d_d
= 0
\\[6pt]
& g_g^{\min} \le g_g \le g_g^{\max}
\quad \forall g \in \mathcal{G}
\\[6pt]
& 0 \le d_d \le d_d^{\text{nom}}
\quad \forall d \in \mathcal{D}
\\[6pt]
& -F_l^{\max}
\le
\sum_{s \in \mathcal{N}} PTDF_{l,s}
\left(
\sum_{g: n(g)=s} g_g
-
\sum_{d: n(d)=s} d_d
\right)
\le
F_l^{\max}
\quad \forall l \in \mathcal{L}
\\[6pt]
& -F_l^{\max}
\le
\sum_{s \in \mathcal{N}} PTDF_{l,s}^{(k)}
\left(
\sum_{g: n(g)=s} g_g
-
\sum_{d: n(d)=s} d_d
\right)
\le
F_l^{\max}
\quad \forall l \in \mathcal{L}, \forall k \in \mathcal{K}
\end{align}

\end{document}